Global adjoint tomography -- Model GLAD-M25

### 2018-10-28 main.tex
Updates from ShareLaTeX

The deployment of USArray has dramatically increased data overage across the
continental United States.
In this section, we compare GLAD-M25 in this region with
with two global models, S362ANI$+$M and S40RTS, and two regional models,
US22~\citep{zhu2017radial} and SL2013NA~\citep{schaeffer2014imaging}.
US22 is a radially anisotropic model based on adjoint tomography using
USArray data from 180 regional earthquakes using 15--50~s
body waves and 25--150~s surface waves.
SL2013NA is a global upper-mantle shear wavespeed model based on multimode
surface waveform tomography primarily focused on North America.

@@

At a depth of 100~km, we observed that the edges of the craton are sharper in
GLAD-M15 than in the other two global models,
and quite similar in shape and sharpness to the two regional models.
We can clearly see impressions of the Snake River Plain (with Yellowstone)
and the Raton hotspot, which also have distinct slow expressions at depths of 200~km.
The core of the craton exhibits fast wavespeeds at depths reaching 300~km Northwest of
Hudson Bay.
In the regional models the craton has basically faded away at these depths.
The Wyoming and Medicine Hat Blocks show a very strong and thick lithospheric
root from shallow depths to $\sim$400~km, similar to US22 but
difficult to discern in SL2013NA.
Similar observations can be made for the Yavapai and
Mazatzal Blocks, which define the southern boundary of the large craton.

@@

Fig.~\ref{fig:europe-vs} shows isotropic shear wavespeed perturbations beneath Europe.
In addition to our model GLAD-M25, S362ANI+M and S40RTS, we included a regional
tomography model EU60\citep{zhu2015seismic} into our comparison. The EU60 uses
adjoint methods to iteratively inverse the crust and upper mantle structure
beneath the European continent and North Atlantic Ocean, by assimilating
190 earthquakes and 745 seismic station.

Europe is featured by the Trans-European Suture Zone(TESZ), separating the
Precambrian East European Craton(EEC) in the northeast and the Phanerozoic
Europe in the southwest.  The TESZ is very consistent across all the models,
even though is much sharper in GLAD-M25 and EU60.

In the north, we observed the high velocity anomaly beneath Central Graben in the
North Sea, and clearly separated from the EEC bounded at the western north, which is
explained as the lithospheric drip with delamination, as explained in
\cite{zhu2015seismic}.

In the south at the Mediterranean region, the tectonic settings is complex with the
existence of several regional plates. In the west of Mediterranean ocean, there is
Appennines-Maghrebides subductions, with the subduction retreat direction from South
(older, >30Ma)to North(younger, present), with its western border of back-arc basin
following along the coastline of East Spain and South France
\citep{Giordano2017Magmatic}. In our model, this is shown as fast velocity starting
from 400km and extend all the way to the mantle transition zones at 600km beneath the
Mediterranean Ocean, nicely following its west and east boundaries. At shallower depth
such as 100km, we observe low velocity structures, which is explained as the
back-arc extension associated with the roll-back. Compared to S362ANI+M and S40RTS,
our model shows much sharper features for the tectonic structures mentioned above.

Moving further to the east, at about 400km, there is clear fast anomaly lie
beneath the Dinarides Mountains, separating the Adria plate and Pannonian Basin,
indicating the subduction of Adria plate into the Euroasian plate.
This anomaly get sharper at 600km. The southern
part of it interacts with the Hellenic Arc subductions, which penetrate the mantle
transition zones and extend below 1,000 km in our model. We will discuss the Hellenic
Arc in details in the later sections. The Pannonian Basin is featured by low velocity
anomaly at the depth from 100 to 250km, then gradually transferred to high velocity
due to the subduction of Adria plate. The Anatolia plate is shown as slow velocity
anomaly in our model, sharpened from two other global models.

In the Central Europe, there is Cenozoic Ridge System, featured by the low velocity
anomaly, starting from Massif Central and extending to Eifel. Both the EU60 and
GLAD-M25 found slow anomaly lying beneath the northern par of Rhine Graben, explained
as the reservoir that fuels the Eifel hotspot, from 100km to 250km.

In the north of Atlantic Ocean, our model strengthened the slow anomaly near the
Atlantic Ocean Ridge, especially at the Iceland and Azores. The Iceland, becomes
much strong in our model in the upper mantle from 250km to 400km, compared to
S362ANI+M, extending all the way to the mantle-transition zone. On the other hand,
Azores, is a much shallower hotspot, with the slow anomaly extend at about 250km.

@@

Fig.\ref{fig:southamerica-vp} shows the $V_p$ velocity in our model, together with 3
global models, S362ANI+M, LLNL-G3Dv3\citep{}, GAP-P4\citep{fukao2013subducted},
while the latter two are global $V_p$ models.

At 100km depth, the northern part of South America is covered by large area
of lithosphere craton, showing fast wave anomaly except GAP-P4. The fast anomaly
stays strong in LLNL-G3Dv3 while for GLAD-M25
and S362ANI+M it disappear at such depth.
At 300km, we are see a sharpened fast anomaly in the south subduction area
in GLAD-M25 compared to S372ANI+M. However, the shape and strength are very close
between GLAD-M25 and LLNL-G3Dv3 models. Moving to even larger depth, the features of
subductions becomes divergent between the north Peruvian slab and south Chilean slab.
At 600km, we observed a large chunk of fast anomaly in the Chilean slab,
indicating the slab is trapped and stagnant in the mantle transition zone. However,
the Peruvian slabs has penetrated the mantle transition zone and extend till 1000,
reaching the uppermost part of lower mantle, which is consistent with GAP-P4
\citep{fukao2013subducted}.

>>>>

The deployment of USArray has dramatically increased data overage across the
continental United States.
In this section, we compare GLAD-M25 in this region 
with two global models, S362ANI$+$M and S40RTS, and two regional models,
US22~\citep{zhu2017radial} and SL2013NA~\citep{schaeffer2014imaging}.
US22 is a radially anisotropic model based on adjoint tomography using
USArray data from 180 regional earthquakes using 15--50~s
body waves and 25--150~s surface waves.
SL2013NA is a global upper-mantle shear wavespeed model based on multimode
surface waveform tomography primarily focused on North America.

@@

At a depth of 100~km, we observe that the edges of the craton are sharper in
GLAD-M15 than in the other two global models,
and quite similar in shape and sharpness to the two regional models.
We can clearly see impressions of the Snake River Plain (with Yellowstone)
and the Raton hotspot, which also have distinct slow expressions at depths of 200~km.
The core of the craton exhibits fast wavespeeds at depths reaching 300~km Northwest of
Hudson Bay.
In the regional models the craton has basically faded away at these depths.
The Wyoming and Medicine Hat Blocks show a very strong and thick lithospheric
root from shallow depths to $\sim$400~km, similar to US22 but
difficult to discern in SL2013NA.
Similar observations can be made for the Yavapai and
Mazatzal Blocks, which define the southern boundary of the large craton.

@@

Fig.~\ref{fig:europe-vs} shows isotropic shear wavespeed perturbations beneath Europe.
In addition to models GLAD-M25, S362ANI$+$M, and S40RTS, we included regional
model EU60~\citep{zhu2015seismic} for comparison.
EU60 is the result of adjoint tomography of the crust and upper mantle
beneath the European continent and the North Atlantic based on the assimilation of data from
190 earthquakes recorded by 745 seismographic station.

A defining tectonic feature of Europe is the Trans-European Suture Zone~(TESZ),
also known as the Tornquist-Teisseyre Zone,
separating the  Precambrian East European Craton~(EEC) in the Northeast from Phanerozoic Europe in the Southwest.
The TESZ is very consistent across all models,
even though is much sharper in GLAD-M25 and EU60.

At depths of 100---250~km we observe high wavespeeds beneath the Central Graben in the North Sea, which is clearly separated from the EEC.
The tectonic history of the Mediterranean is complex,
involving several regional plates and multiple past and present subduction zones.
We see slab ponding in the transition zone beneath the western Mediterranean associated with slab roll-back of the Appennines-Maghrebides subduction zone~\citep{Giordano2017Magmatic}.
At shallower depths, we see low wavespeeds along its western border,
which are interpreted as back-arc extension associated with the roll-back. Compared to S362ANI$+$M and S40RTS,
our model shows much sharper features for these tectonic structures.

In the Eastern Mediterranean, there is a distinct fast anomaly beneath the 
Dinarides Mountains, separating the Adria plate from the Pannonian Basin,
indicating subduction of the Adria plate beneath the Eurasian plate.
This anomaly gets sharper at 600~km depth.
Its southern part interacts with the Hellenic slab, which penetrates the mantle transition zone and extends below 1,000~km in our model.
We discuss the Hellenic Arc further in Section~\ref{section:slabs}.
The Pannonian Basin is characterized by a low wavespeed
anomaly at depths from 100~km to 250~km.
In this depth range we also see slow wavespeeds beneath Anatolia,
stronger than in the two other global models.

In Central Europe we note the Cenozoic Ridge System, characterized by low  wavespeeds from the Massif Central to the Eifel. Both EU60 and
GLAD-M25 exhibit a slow anomaly beneath the northern part of the Rhine Graben, which is interpreted as the reservoir that fuels the Eifel hotspot.

Beneath the Atlantic Ocean, our model strengthens slow anomalies beneath the ridges, especially underneath Iceland and the Azores.
Compared to S362ANI$+$M, Iceland becomes
much stronger in the upper mantle at depths from 250~km to 400~km, extending all the way to the mantle-transition zone.
The Azores represent a much shallower hotspot, with a maximum depth of about 250~km.

@@

Fig.\ref{fig:southamerica-vp} shows the $V_p$ velocity in our model, together with 3
global models, S362ANI$+$M, LLNL-G3Dv3\citep{}, GAP-P4\citep{fukao2013subducted},
while the latter two are global $V_p$ models.

At 100km depth, the northern part of South America is covered by large area
of lithosphere craton, showing fast wave anomaly except GAP-P4. The fast anomaly
stays strong in LLNL-G3Dv3 while for GLAD-M25
and S362ANI$+$M it disappear at such depth.
At 300km, we are see a sharpened fast anomaly in the south subduction area
in GLAD-M25 compared to S372ANI+M. However, the shape and strength are very close
between GLAD-M25 and LLNL-G3Dv3 models. Moving to even larger depth, the features of
subductions becomes divergent between the north Peruvian slab and south Chilean slab.
At 600km, we observed a large chunk of fast anomaly in the Chilean slab,
indicating the slab is trapped and stagnant in the mantle transition zone. However,
the Peruvian slabs has penetrated the mantle transition zone and extend till 1000,
reaching the uppermost part of lower mantle, which is consistent with GAP-P4
\citep{fukao2013subducted}.
